We present a reduction semantics for lqCCS, consistent with the labelled semantic for qCCS presented in \cite{fengBisimulationQuantumProcesses2012, dengOpenBisimulationQuantum2012}. A reduction semantic does not make any assumption on the observable properties of the system (like the labels of a transition), and so is better suited to explore and compare different notion of behavioural equivalence.

Besides, as explained in Chapter \ref{chapter3}, in a labelled transition system a quantum input transition like $\xrightarrow{c?q}$ requires the value of qubit $q$ to be already present in the state. This is an atypical assumption, as a labeled transition usually model the communication with an unknown external environment, but in this case requires at least some partial knowledge of the environment. In a reduction system there are no such labelled transitions, so a process communicates only with other processes, on which we have total information.

Our semantics defines a probabilistic reduction system $\langle S, \rightarrow \rangle$, where \begin{itemize}
\item $S$ is a set of \textit{configurations}, of the form $\confw{\rho, P}$ (like in qCCS).
\item $\rightarrow \subseteq S \times \distr(S)$ is the probabilistic transition relation, corresponding to the $\xrightarrow{\tau}$ transition in qCCS \cite{fengBisimulationQuantumProcesses2012, dengOpenBisimulationQuantum2012}.
\end{itemize}

We assume a fixed set  $QN = {q_1, q_2, \ldots q_n}$ of quantum names, where each name $q_i$ refers to a unique qubit with state space $\calH_i$. We denote as $\calH_{QN}$ the $2^n$-dimensional Hilbert space $\bigotimes_{i=1}^n \calH_i$, and so any state 
$\rho \in \mathcal{D}(\calH_{QN})$ associates each name with a value.

We also assume a fixed typing contest $\Gamma_c = \set{c_i: \widehat{T_i}}_i$, containing typing assumptions for a finite set of classical and quantum channels.

\begin{definition}
Let $P$ be a process and $\rho \in \mathcal{D}(\calH_{QN})$ an arbitrary density operator. We say that a configuration $\confw{\rho, P}$ is well typed, given a set of quantum names $QN$ and a set of typed channels $\Gamma_c$, if $\Gamma_c; \Sigma \vdash P$ for some $\Sigma \subseteq QN$.

%  Let $I$ be an arbitrary index set. $\Gamma; \Sigma \vdash \boxplus_{i \in I} \langle \rho_i, P_i \rangle$ iff for each $i \in I$ such that $\rho_i \neq \mathbf{0}$, then $\Gamma; \Sigma \vdash \langle \rho_i, P_i \rangle$.
\end{definition}

Notice that the context $\Gamma_c$ contains assignments only for channels, not for classical variables. This means that in well typed configuration, $P$ does not contain any free classical variable, and all the free quantum variables are references to qubits in the configuration.

From now on, we will consider only well typed configurations.

\subsection{Reduction System}
In order to define the reduction transition, we first need to introduce a semantic for expressions and a structural congruence relation on processes, like in \cite{gayCommunicatingQuantumProcesses2005}.

We consider as a \textit{value} any expressions $n \in \mathbb{N},\  b \in \mathbb{B},\  x \in  \text{Var}$, and use $v$ as a metavariable for them.

In Figure \ref{big_step_exp}, we define a big step semantic for classical and quantum expression in the usual way. We write $e \Downarrow v$ to indicate that the expression $e$ evaluates to value $v$. Recall that the only quantum expression admitted by the type system are quantum variables.

\begin{figure}[h!]
\begin{gather*}
\infer[\mbox{\footnotesize\scshape Var}]{x \Downarrow x}{} \qquad
    \infer[\mbox{\footnotesize\scshape Nat}]{n \Downarrow n}{} \qquad
    \infer[\mbox{\footnotesize\scshape Bool}]{b \Downarrow b}{}  \qquad
\\[0.3cm]
    \infer[\mbox{\footnotesize\scshape Or}]{(e_1 \lor e_2) \Downarrow b}{e_1 \Downarrow b_1 & e_2 \Downarrow b_2 & b = b_1 \lor b_2} \qquad
    \infer[\mbox{\footnotesize\scshape Neg}]{\neg e \Downarrow b}{e \Downarrow b_1 & b = \neg b_1} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape Leq}]{(e_1 \leq e_2) \Downarrow b}{e_1 \Downarrow n_1 & e_2 \Downarrow n_2 & b = n_1 \leq n_2}
\end{gather*}
\caption{Big step semantic for Linear qCCS expressions}
\label{big_step_exp}
\end{figure}

We define as the \textit{structural congruence relation} $\equiv$ the smallest equivalence relation that satisfies the axioms in Figure \ref{str_cong}. We employed the usual axioms of \cite{milnerFunctionsProcesses1990} for parallel composition, summation and reduction. We also add axioms to evaluate classical expressions and resolve \textbf{If Then Else} constructs.

\begin{figure}[h!]
\begin{gather*}
    \infer[\mbox{\footnotesize\scshape ParNil}]{P \parallel nil \equiv P}{} \qquad
    \infer[\mbox{\footnotesize\scshape ParComm}]{P \parallel Q \equiv Q \parallel P}{} \qquad
    \infer[\mbox{\footnotesize\scshape ParAssoc}]{P \parallel (Q \parallel R) \equiv (P \parallel Q) \parallel R}{} 
    \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape SumNil}]{M + nil \equiv M}{} \qquad
    \infer[\mbox{\footnotesize\scshape SumComm}]{M + N \equiv N + M}{} \qquad
    \infer[\mbox{\footnotesize\scshape SumAssoc}]{M + (N + O) \equiv (M + N) + O}{} 
    \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape RestrOrd}]{P \setminus c \setminus d \equiv P \setminus d \setminus c}{} \qquad 
    \infer[\mbox{\footnotesize\scshape RestrNil}]{nil \setminus c \equiv nil}{} \qquad 
    \infer[\mbox{\footnotesize\scshape RestrPar}]{(P \parallel Q) \setminus c \equiv P \parallel (Q \setminus c)}{c \not\in fc(P)} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape TrueGuard}]{\ITE{tt}{P}{Q} \equiv P}{} \qquad
    \infer[\mbox{\footnotesize\scshape FalseGuard}]{\ITE{ff}{P}{Q} \equiv Q}{} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape ValExpr}]{ P \equiv P[\sfrac{v}{e}] }{e \Downarrow v} 
	\end{gather*}	
\caption{Structural congruence for Linear qCCS}
\label{str_cong}
\end{figure}

We can now define the transition relation $\rightarrow$, presented in Figure \ref{reduction}. As usual, we write $\confw{\rho, P} \rightarrow \Delta$ to intend $(\confw{\rho, P}, \Delta) \in \rightarrow$. To lighten the notation, we will also write $\confw{\rho, P} \rightarrow \confw{\rho', P'}$ instead of $\confw{\rho, P} \rightarrow \overline{\confw{\rho', P'}}$.

This transition relation cannot be composed with itself, since it is a relation from configurations to distributions. As in \cite{fengBisimulationQuantumProcesses2012}, we can \textit{lift} the relation $\rightarrow$ into $\slift{\rightarrow}$, that is a relation from distributions to distributions, as described in Section \ref{pLTS}. This is useful to talk about reachability and temporal logics, but will not be used in the definitions of strong bisimulations in the next sections.

\begin{figure}[h!]
  \begin{gather*}
    \infer[\mbox{\footnotesize\scshape SemTau}]{\langle \rho, \tau . P \rangle \longrightarrow \langle \rho, P \rangle}{} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape SemRename}]{\langle \rho, P \{f\} \rangle \longrightarrow \langle \rho', P' \{f\}\rangle}{\langle \rho, f(P) \rangle \longrightarrow \langle \rho', f(P') \rangle} \qquad
    \infer[\mbox{\footnotesize\scshape SemRestrict}]{\langle \rho, P \setminus L \rangle \longrightarrow \langle \rho, P' \setminus L \rangle}{\langle \rho, P \rangle \longrightarrow \langle \rho', P' \rangle} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape SemQOp}]{\langle \rho, \mathcal{E}(\widetilde{x}) . P \rangle \longrightarrow \langle \mathcal{E}_{\widetilde{x}}(\rho), P \rangle}{} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape SemQMeas}]{\langle \rho, M(\widetilde{x} \rhd y) . P \rangle \longrightarrow \sum_{m = 0}^{2^{|\widetilde{x}|}} p_m(\rho) \left\langle \frac{1}{p_m}\rho_m , P[\sfrac{m}{y}] \right\rangle}{\rho_m = M_m \rho M_m^\dag & p_m = tr(\rho_m)} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape SemPar}]{\langle \rho, P \parallel R \rangle \longrightarrow \langle \rho', P' \parallel R \rangle}{\langle \rho, P \rangle \longrightarrow \langle \rho', P' \rangle} \qquad
    \infer[\mbox{\footnotesize\scshape SemSum}]{\langle \rho, P + R \rangle \longrightarrow \langle \rho', P' \rangle}{\langle \rho, P \rangle \longrightarrow \langle \rho', P' \rangle} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape SemReduce}]{\langle \rho, c!v \parallel c?x . P \rangle \longrightarrow \langle \rho, P[\sfrac{v}{x}] \rangle}{} \\[0.3cm]
    \infer[\mbox{\footnotesize\scshape SemCongr}]{\langle \rho, P \rangle \longrightarrow \langle \rho', P' \rangle}
    {P \equiv Q & \confw{\rho, Q} \rightarrow \confw{\rho', Q'} & Q' \equiv P'}
  \end{gather*}
\caption{Reduction system for Linear qCCS}
\label{reduction}
\end{figure}
  
\begin{example}
We can formalize in lqCCS the famous \textit{quantum teleportation} algorithm:
\begin{align*}
  \proc{A} &\Coloneqq \text{in}_a?x.\text{CNOT}(q_0, x).\text{H}(q_0).M(q_0, x \rhd n).(\text{m}_ab!n \parallel discard(q_0, x)\\
  \proc{B} &\Coloneqq \text{in}_b?x.\text{m}_{ab}?n.
     \\ & \ite{n = 0}{I(x).\text{out}_b!x\\&}
    {\ite{n = 1}{X(x).\text{out}_b!x\\&}
    		{\ite{n = 2}{Z(x).\text{out}_b!x\\&}{Y(x).\text{out}_b!x}}
    }\\
  \proc{S} &\Coloneqq Set_{\Phi^+}(q_1, q_2).(\text{in}_a!q_1 \parallel \text{in}_b!q_2) \\
  \proc{Tel} &\Coloneqq (A \parallel B \parallel S) \setminus \Set{\text{in}_a, \text{in}_b, \text{m}_{ab} }
\end{align*}
Which is well typed under the context 
$\Gamma = \set{\text{out}_b : \hat{\mathcal{Q}}}$, $\Sigma = \set{q_0, q_1, q_2}$.
\end{example}